Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-13 analysis/paper/paper.Rmd
add script file for metric data
update the metric code, and edit the background

The predominant artefact found is locally made ceramics, which distributed throughout the sequence across the site. More than 550 thousands of sherds were excavated from the upper component at Kiwulan, and around 1,200 vessels were reconstructed. It is interesting to note that there are only two forms of vessels recovered, one is cooking pot and another one is steamer made of two cooking pots stacked together with a filter clay layer in between. Those pots show high consistency in shape. They are globular body with short neck and wide mouth. The exterior surface below the neck is treated with a wide variety of impressed geometric motifs. They were believed used for cooking because of the evidence of charred residues and carbon deposits commonly observed on the interior and soot on the exterior. Moreover, there is no other utilitarian earthenware ceramics found at this site suggests this type of pot was mainly used for daily cooking. Most pots are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes. Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand.

We examine 291 reconstructed pots recovered from the 40 sampling units. The layer from 1 to 6 could be divided into 3 time periods. Layer 5 and 6 represent pre-contact period, layer 4 represents contact period, and layer1, 2, and 3 represent post-contact period. The amount of pots for each layer shows below. Although most pots are not complete, the thickness from rim, neck, to body can be measured. I have also measured the diameter of the rim, neck, and body. For those pots that are incomplete, the diameter is measured by its curvature. Since the height is incomplete for most pots, this preliminary analysis focuses more on the possible change in thickness and diameter of pot for different parts, and their ratio over time.

@@

More standardized pottery shape without controlling the clay paste or technology suggests the shape as a preferred characteristic in pottery production during the Chinese period. However, the standardization is not reflected by the metric attributes, specifically the rim, neck, and body thickness. There is higher variation in thickness attributes compared to diameter attributes, indicating shape is the only attribute that presents homogeneous. In addition to the emergence of specialized groups indicative of changes in social organization, another explanation is that producing homogeneous pottery is an act of resistance to the Chinese immigrant by emphasizing the practice of traditional pottery production. Among the pottery attributes, shape is the most  obvious and visible trait, which can be used as an expression of social boundary or group identity [cf. @Raux2015].

>>>>

The predominant artefact found is locally made ceramics, which distributed throughout the sequence across the site. More than 550 thousands of sherds were excavated from the upper component, and around 1,200 vessels were reconstructed. It is interesting to note that there are only two forms of local vessels, one is cooking pot and another one is steamer made of two cooking pots stacked together with a filter clay layer in between. Those pots show high consistency in shape. They are globular body with short neck and wide mouth. The exterior surface below the neck is treated with a wide variety of impressed geometric motifs. They were believed used for cooking because of the evidence of charred residues and carbon deposits commonly observed on the interior and soot on the exterior. Moreover, a lack of evidence of other utilitarian earthenware ceramics found at this site suggests the globular pot was mainly used for daily cooking. They are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes. Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand.

The metric attributes for 362 reconstructed pots (pre-European = 153, European = 173, Chinese = 36) from the 40 sampling units were measure and compared using coefficient of variation (CV) with a significance test for the equality of CVs. The attributes include thickness of rim, neck, and body, and diameter of rim, neck, and body. Two measurements for thickness and diameter for each pot were took to obtain the average value. To understand more about the pottery shape, we also compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter. The results show that there is a significant difference (P-value < 0.01) in the variation of two metric attributes, the rim diameter and the neck diameter across three phases. The CV values show that the variation of rim diameter is small (6.8%) before European contact, increases after the European presence (9.6%), and decreases after the Chinese presence (6.9%). For the neck diameter, the CV values indicate smaller variation after the European period (7.7%) and the Chinese period (7.1%) compared to pre-European period(10.6%).

@@

More standardized pottery shape without controlling the clay paste or technology suggests the shape as a preferred characteristic in pottery production during the Chinese period. However, the standardization is not reflected by the metric attributes, specifically the rim, neck, and body thickness. There is higher variation in thickness attributes compared to diameter attributes, indicating shape is the only attribute that presents homogeneous. In addition to the emergence of specialized groups indicative of changes in social organization, another explanation is that producing homogeneous pottery is an act of resistance to the Chinese immigrant by emphasizing the practice of traditional pottery production. Among the pottery attributes, shape is the most  obvious and visible trait, which can be used as an expression of social boundary or group identity [cf. @Roux2015].
